---
title: "Lab 07 — t-tests and ANOVA"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

GRADE GM: 10/10 - Very strong submission! Clean code, thoughtful interpretations, excellent visualizations, and a solid reflection. Excellent work!

```{r setup, echo=FALSE, message=FALSE, warning=T}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab extends hypothesis testing from **claims about one mean** to:

- **one-sample t-tests**
- **two-sample t-tests**
- **paired t-tests**
- **ANOVA**

You will use the **US-AMS simplified AmeriFlux dataset** again and compare means across different time-based groups.

The core logic stays the same:

> estimate → uncertainty → test statistic → p-value

This lab emphasizes both **manual understanding** and **automated testing in R**.

---

# Learning goals

By the end of this lab, you should be able to:

- Compute and interpret a **one-sample t-test**
- Distinguish between **two-sample** and **paired** t-tests
- Compute and interpret **degrees of freedom**
- Explain the structure of an **ANOVA**
- Compare manual calculations to `t.test()` and `aov()`
- Use visualizations to support statistical interpretation
- Produce a fully reproducible R Markdown analysis

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=T}
# GOAL: Read the dataset and keep 2022–2023.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local %in% c(2022, 2023))

glimpse(df)
```

## Derive time columns if needed

```{r derive_time, echo=TRUE, eval=T}
# GOAL: Create date / month columns if they do not already exist.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE)
  )
```

---

# Part I — One-sample t-test

## 1) Choose a day and a variable

Select:
- one **year**
- one **numeric variable**

Examples:
- `ta`
- `swc`
- `le`
- `fc`

```{r choose_one_sample, echo=TRUE, eval=T}
target_year <- 2022
target_var  <- "ta"

sample_data <- df %>%
  filter(year_local == target_year) %>% 
  slice_sample(n = 5000)

x <- sample_data[[target_var]]

mean(x, na.rm = TRUE); sd(x, na.rm = TRUE); length(x)
```

> What variable and day did you choose, and why?
> *I chose Air Temperature (ta) because I used the same variable in previous labs, and now I want to compare it with the 2022 data.*

---

## 2) State a null hypothesis

Write a claim about the mean of your chosen variable on that day.

Examples:
- “The mean temperature on this day is 15 °C.”
- “The mean latent heat flux on this day is 100 W m$^{-2}$.”

Fill in:

> $H_0$: $\mu = 11$

> $H_A$: $\mu \neq 11$

---

## 3) Compute the one-sample t-statistic manually

Use:

$t = (\bar{x} - \mu_0) / (s / \sqrt{n})$

```{r one_sample_manual, echo=TRUE, eval=T}
mu0 <- 11

xbar <- mean(x, na.rm = TRUE)
s <- sd(x, na.rm = TRUE)
n <- sum(!is.na(x))

t_stat <- (xbar - mu0) / (s / sqrt(n))
t_stat
```

---

## 4) Compute the p-value and degrees of freedom

```{r one_sample_pvalue, echo=TRUE, eval=T}
df_t <- n - 1

p_value <- 2 * pt(-abs(t_stat), df = df_t)

tibble(
  xbar = xbar,
  s = s,
  n = n,
  df_t = df_t,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the p-value suggest about your null hypothesis?
> *The p-value is extremely small, so I can reject the null hypothesis. This suggests that the true mean ta is statistically significantly different from my hypothesized value of 11.*

> GM: Excellent — Clear rejection of null and correct interpretation.

---

## 5) Compare with `t.test()`

```{r one_sample_ttest, echo=TRUE, eval=T}
t.test(x, mu = mu0)
```

> Does the `t.test()` output agree with your manual calculation? If not, why might that be?
> *Yes, they agree. The manual calculation and the automatic output both resulted in the same t-statistic and degrees of freedom.*

---

# Part II — Two-sample t-test

## 6) Compare the same variable across two day ranges

Choose:
- the **same variable**
- two different **day ranges in 2023**

```{r choose_two_sample, echo=TRUE, eval=T}
x1 <- df %>%
  filter(year_local == 2023, doy > 95 & doy < 97) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 194 & doy < 196) %>%
  pull(target_var)

mean(x1, na.rm = TRUE); sd(x1, na.rm = TRUE); length(x1)
mean(x2, na.rm = TRUE); sd(x2, na.rm = TRUE); length(x2)
```

> Why did you choose these two days?
> *To compare seasonal temperature differences between spring and summer, so I can see if there is a statistically significant shift.*

## 7) Summarize the difference in means

```{r two_sample_summary, echo=TRUE, eval=T}
xbar1 <- mean(x1, na.rm = TRUE)
xbar2 <- mean(x2, na.rm = TRUE)

s1 <- sd(x1, na.rm = TRUE)
s2 <- sd(x2, na.rm = TRUE)

n1 <- sum(!is.na(x1))
n2 <- sum(!is.na(x2))

mean_diff <- xbar1 - xbar2

tibble(
  xbar1 = xbar1,
  xbar2 = xbar2,
  mean_diff = mean_diff
)
```

---

## 8) Compute the two-sample t-statistic manually (Welch)

```{r two_sample_manual, echo=TRUE, eval=T}
se_diff <- sqrt((s1^2 / n1) + (s2^2 / n2))

t_stat <- mean_diff / se_diff

df_welch <- ((s1^2 / n1 + s2^2 / n2)^2) /
  (((s1^2 / n1)^2 / (n1 - 1)) + ((s2^2 / n2)^2 / (n2 - 1)))

p_value <- 2 * pt(-abs(t_stat), df = df_welch)

tibble(
  mean_diff = mean_diff,
  se_diff = se_diff,
  t_stat = t_stat,
  df_welch = df_welch,
  p_value = p_value
)
```

> What does the sign of `mean_diff` mean? What does the p-value mean?
> *The negative sign means that the average spring temperature is lower than the average summer temperature. The extremely small p-value means that this difference is statistically significant.*

---

## 9) Compare with `t.test()`

```{r two_sample_ttest, echo=TRUE, eval=T}
t.test(x1, x2)
```

> Does the automated t-test agree with your manual result?
> *Yes, they agree.*

---

## 10) Visualize the two groups

```{r two_sample_plot, echo=TRUE, eval=T}
day_range1 <- "95_97"
day_range2 <- "194_196"
plot_df <- bind_rows(
  tibble(group = paste0("DOY_", day_range1), value = x1),
  tibble(group = paste0("DOY_", day_range2), value = x2)
)

ggplot(plot_df, aes(x = group, y = value, fill = group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = target_var,
    title = "Two-sample comparison"
  )
```

> Does the plot match your statistical conclusion?
> *Yes. The boxplot shows that the summer box is higher on the y-axis than the spring box, with almost no overlap. This visual separation confirms the significant difference detected by the t-test.*

> GM: Excellent — Strong tie between plot and statistical conclusion.

---

# Part III — Paired t-test

## 11) Compare the same days across two years

Choose one DOY and compare **2022 vs 2023** for the same variable.

```{r paired_setup, echo=TRUE, eval=T}
x1 <- df %>%
  filter(year_local == 2022, doy > 145 & doy < 165) %>%
  arrange(date) %>%
  pull(target_var)

x2 <- df %>%
  filter(year_local == 2023, doy > 145 & doy < 165) %>%
  arrange(date) %>%
  pull(target_var)

length(x1); length(x2)
```

> Why is this a paired comparison rather than an independent two-sample comparison?
> *It is paired because the observations are at the exact same time period measured across two different years.*

---

## 12) Compute paired differences manually

```{r paired_manual, echo=TRUE, eval=T}
d <- x1 - x2

dbar <- mean(d, na.rm = TRUE)
sd_d <- sd(d, na.rm = TRUE)
n <- sum(!is.na(d))

t_stat <- dbar / (sd_d / sqrt(n))
p_value <- 2 * pt(-abs(t_stat), df = n - 1)

tibble(
  dbar = dbar,
  sd_d = sd_d,
  n = n,
  t_stat = t_stat,
  p_value = p_value
)
```

> What does the sign of `dbar` mean?
> *This means that the ta on those specific days were slightly warmer on average in 2022 compared to 2023.*

---

## 13) Compare with `t.test(..., paired = TRUE)`

```{r paired_ttest, echo=TRUE, eval=T}
t.test(x1, x2, paired = TRUE)
```

> How does the paired t-test differ conceptually from the two-sample t-test?
> *An independent two-sample t-test compares the overall means of two completely unrelated groups. A paired t-test calculates the difference between paired observations.*

> GM: Excellent — Clean and accurate conceptual distinction.

---

# Part IV — ANOVA

## 14) Create a categorical grouping variable

Group `ta` into broad seasonal categories:

- Summer
- Winter
- Other

```{r make_groups_anova, echo=TRUE, eval=T}
df_anova <- df %>%
  filter(!is.na(ta)) %>%
  mutate(
    month_group = case_when(
      month %in% c("Jun", "Jul", "Aug") ~ "Summer",
      month %in% c("Dec", "Jan", "Feb") ~ "Winter",
      TRUE ~ "Other"
    )
  )

df_anova %>% count(month_group)
```

---

## 15) Visualize the groups

```{r plot_anova_groups, echo=TRUE, eval=T}
ggplot(df_anova, aes(x = month_group, y = ta, fill = month_group)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = "ta",
    title = "Air temperature by seasonal group"
  )
```

> Do the groups look different enough that ANOVA might detect a difference?
> *Yes. The boxplot enough separation between groups. Because the boxes do not overlap much, this implies the means are far enough apart for ANOVA to detect a difference.*

---

## 16) Compute ANOVA components manually

```{r anova_manual_components, echo=TRUE, eval=T}
grand_mean <- mean(df_anova$ta, na.rm = TRUE)

group_stats <- df_anova %>%
  group_by(month_group) %>%
  summarise(
    group_mean = mean(ta, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ss_between <- sum(group_stats$n * (group_stats$group_mean - grand_mean)^2)

df_anova2 <- df_anova %>%
  left_join(group_stats, by = "month_group")

ss_within <- sum((df_anova2$ta - df_anova2$group_mean)^2)

k <- n_distinct(df_anova$month_group)
n_total <- nrow(df_anova)

df_between <- k - 1
df_within <- n_total - k

ms_between <- ss_between / df_between
ms_within <- ss_within / df_within

f_stat <- ms_between / ms_within
p_value <- pf(f_stat, df1 = df_between, df2 = df_within, lower.tail = FALSE)

tibble(
  grand_mean = grand_mean,
  ss_between = ss_between,
  ss_within = ss_within,
  df_between = df_between,
  df_within = df_within,
  f_stat = f_stat,
  p_value = p_value
)
```

> In your own words, what do `ss_between` and `ss_within` represent?
> *ss_between: Represents the variation between the different groups.*
> *ss_within: Represents the normal variation within each group.*

---

## 17) Compare with `aov()`

```{r run_anova, echo=TRUE, eval=T}
model <- aov(ta ~ month_group, data = df_anova)
summary(model)
```

> What does the ANOVA p-value tell you? Does it tell you which groups differ?
> *This p-value tells me that at least one of the seasonal group mean is significantly different from the others.*

---

## 18) Final reflection

Answer in **5–7 sentences**:

- How is a one-sample t-test different from a two-sample t-test?
- Why is a paired t-test special?
- How is ANOVA a generalization of the t-test?
- What stayed the same across all these tests?

> *A one-sample t-test compares a single group mean to a hypothesized value, whereas a two-sample t-test compares the means of two entirely independent groups. A paired t-test is unique because it analyzes dependent data. ANOVA serves as a generalization of the t-test, allowing us to compare means across three or more groups simultaneously without inflating our error rate. The logic stayed the exact same across all these tests by calculating a test statistic and then comparing the estimated effect size against the data uncertainty.*

> GM: Excellent — Very good summary of the key differences and common logic.

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.
